---
title: "Trophic Networks Robustness workflow"
author: 
  -   "Ceres Barros"
  -   "Gabriel Dansereau"
  -   "João Braga"
date: August 21 2024
output: 
  rmdformats::downcute:    
    toc_depth: 3
    toc_float: true
    code_folding: show
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(cache = 2,
                      warning = FALSE,
                      message = FALSE,
                      results = "hide",
                      tidy = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
options(repos = "https://cran.rstudio.com/")
```

```{r, eval = isTRUE(as.logical(Sys.getenv("CI"))), echo = FALSE, cache = FALSE}
## MacOS systems having lots of issues with pkg install here, so install dealt with by
## r-lib/actions/setup-r-dependencies@v2 using DESCRIPTION

options(repos = c("predictiveecology.r-universe.dev", getOption("repos")))

if (!"Require" %in% installed.packages())
  install.packages("Require")

Require::Require(c(
  "cheddar",
  "colorspace",
  "data.table",
  "dplyr",
  "exactextractr",
  "foreign",
  "future",
  "future.apply",
  "ggExtra",
  "ggplot2",
  "ggpubr",
  "patchwork",
  "reproducible",
  "raster",
  "rasterVis",
  "rredlist",
  "RSQLite",
  "sf",
  "CeresBarros/ToolsCB@master (HEAD)"
))

options(future.globals.maxSize = 3e+9,
        reproducible.useNewDigestAlgorithm = 2,
        reproducible.cachePath = "cache/",
        reproducible.useGDAL = FALSE,
        reproducible.rasterRead = "raster::raster",
        nwarnings = 100000000,
        max.print = 100000000)
```

```{r, eval = isTRUE(as.logical(Sys.getenv("CI"))), echo = FALSE}
## GHA: problem with Win zip using prepInputs - ! error -103 with zipfile in unzGetCurrentFileInfo
options(timeout = 600)
utils::download.file(url = "https://zenodo.org/api/records/13345395/files-archive",
                     destfile = "data/13345395.zip",
                     mode = "wb")
unzip("data/13345395.zip", exdir = "data/", overwrite = TRUE, junkpaths = TRUE)
``` 

All data necessary to run this workflow is available in
[Zenodo](https://zenodo.org/doi/10.5281/zenodo.13334864), and
automatically accessed by the code.

## Study area

In @dansereau trophic networks were calculated across European
countries, Turkey, Georgia and European Russia, under baseline
conditions and scenarios of change.

Here, we reduced computational burden by subsetting to a study area
polygon drawn with random borders and placed in central Europe. The user
is free to change this study area as they see fit.

Note that all lakes are removed.

```{r, results = "hold", fig.cap = "Full study area in blue, with (removed) lakes shown in grey. Subset study area in black."}
## 10K grid dbf - pixel IDs 
mask10kID <- prepInputs(url = "https://zenodo.org/api/records/13345395/files-archive",
                        archive = "13345395.zip",
                        targetFile = "reference_grid_10km.img.vat.dbf",
                        destinationPath = "data/",
                        fun = "foreign::read.dbf")
```
